Relative parabolic factorizations of Chevalley groups

### 2015-08-26 bass-kolster.tex
Slight changes in the notation; further reorganization

\textsc{Case $\Phi=\rB_\ell$, $n=2\ell+1$.} Subdivide $v\in \Ums(2\ell+1, I)$ as $v=(v_+, v_0, v_-)\in R^\ell\times R\times R^\ell$.
Denote by $J\leq I$ the ideal spanned by components of $v_-$.
Since $\sr(I/J)\leq \ell$ we can find $c_1,\dots,c_\ell\in I$ such that for $v' = h \cdot v$, $h = \prod_{i=1}^\ell x_{i,0}(c_i) \in \U(\Phi^+, I)$
one has $\bar{v'}_+=(\bar{v'_1},\ldots, \bar{v'_\ell}) \in \Ums(\ell, I/J)$ and, therefore, $(v'_+, v'_-) \in \Ums(2\ell, I)$.
Now the proof can be finished by repeating the argument for the case $\Phi=\rD_l$ (applied to the subset of long roots of $\rB_\ell$).
%(clearly, the maximal root of $\rD_\ell$ maps to the maximal root of $\rB_\ell$ under the natural embedding $\rD_\ell\subseteq\rB_\ell$). 
\end{proof}

It is easy to see that the proof of the above theorem is effective and gives an estimate of the total number of elementary root unipotents involved in the decomposition.
\begin{cor}
In the assumptions and notation of Theorem~\ref{thm:BassKolster} every element of $\G(\Phi,R,I)$ 
can be factored into a product of one element of $\G(\Delta_1,R,I)$ one element of $Z$ and at most $n(\Phi)$ elementary root unipotents $x_\alpha(s)$ of level $I$, where
$n(\rA_\ell)= 4 \ell - 1$, $n(\rB_\ell)= 8 \ell - 5$, $n(\rC_\ell)= 8 \ell - 5$, $n(\rD_\ell)= 8 \ell - 9$. \end{cor}
\begin{proof}
The above estimates can be obtained by a careful analysis of the proof of the previous theorem.
Cases $\Phi=\rA_\ell, \rC_\ell$ are immediate.
In the case $\Phi=\rD_\ell$ the proof of Theorem~\ref{thm:BassKolster} implies that
\begin{multline}\nonumber
\G(\Phi,R,I) =  \U(\Sigma_\ell,I) \cdot X_{\alpha_1}(I) \cdot \U(\Sigma_2^-\cap\Delta_1,I) \cdot X_{-\alpha\ssub{\mathrm{max}}}(I) \cdot Z  \cdot \\ \cdot \U(\Sigma_1^-,I) \cdot \U(\Sigma_1,I) \cdot \G(\Delta_1,R,I).
\end{multline}
We can present an element $g$ of $\U(\Sigma_\ell, I)$ as a product of $g_1 \in \U(\Sigma_{\{1,2\}} \cap \Sigma_\ell)$ and $g_2\in \U(\Delta_{\{1,2\}}\cap \Sigma_\ell)$.
An examination of the extended Dynkin diagram of $\rD_\ell$ implies that $g_2$ either centralizes or normalizes all factors of the above decomposition (except the last one) and therefore can be moved to the right until it is consumed by $\G(\Delta_1)$.
On the other hand, $g_1$ is a product of at most $2\ell-3$ elementary unipotents, while the width of $\U(\Sigma_1^\pm, I)$ and $\U(\Sigma_2^-\cap\Delta_1)$ in elementary unipotents does not exceed $2\ell-2$ and $2\ell-4$, respectively.
Summing up these upper bounds we obtain
$$n(\rD_\ell) = (2\ell-3) + 1 + (2\ell - 4) + 1 + 2\cdot (2\ell - 2) = 8\ell - 9.$$

@@

\begin{proof}[Proof of Theorem~\ref{thm:SL2width}]
Consider the case $\Phi=\rA_\ell$, $\sr(I)\leq 2$. First of all, notice that we can improve the estimate of the number of factors involved in Bass---Kolster decomposition.
Indeed, when performing the first step of the proof of Theorem~\ref{thm:BassKolster} it suffices to make only $2$ additions of $v_{\ell+1}$ (e.\,g. to $v_{1}$ and $v_2$) to make the first $\ell$ entries of $v'$ form a unimodular column.
In particular, $\G(\Phi, R, I)$ can be presented as a product of one element of $\G(\Delta_1, R, I)$, one element of $Z$ and $3\ell+1$ root unipotents $x_\alpha(s)$, $s\in I$.
The latter elements are contained in a product of $3\ell + 1$ copies of $\SL_2(R, I)$. Now the statement of the theorem follows by induction on $\ell$.

The proof in the case $\Phi=\rC_\ell$ is similar (notice that we use the exceptional isomorphism $\SL_2\cong \Sp_2$).
\end{proof}

>>>>

\textsc{Case $\Phi=\rB_\ell$, $n=2\ell+1$.} Subdivide $v\in \Ums(2\ell+1, I)$ as $v=(v_+, v_0, v_-)\in R^\ell\times R\times R^\ell$.
Denote by $J\leq I$ the ideal spanned by components of $v_-$.
Since $\sr(I/J)\leq \ell$ we can find $c_1,\dots,c_\ell\in I$ such that for $v' = h \cdot v$, $h = \prod_{i=1}^\ell x_{i,0}(c_i) \in \U(\Phi^+, I)$
one has $\bar{v'}_+=(\bar{v'_1},\ldots, \bar{v'_\ell}) \in \Ums(\ell, I/J)$ and, therefore, $(v'_+, v'_-) \in \Ums(2\ell, I)$.
Now the proof can be finished by repeating the argument for the case $\Phi=\rD_\ell$ (applied to the subset of long roots of $\rB_\ell$).
%(clearly, the maximal root of $\rD_\ell$ maps to the maximal root of $\rB_\ell$ under the natural embedding $\rD_\ell\subseteq\rB_\ell$). 
\end{proof}

It is easy to see that the proof of the above theorem is effective and gives an estimate of the total number of elementary root unipotents involved in the decomposition.
\begin{cor}
In the assumptions and notation of Theorem~\ref{thm:BassKolster} every element of $\G(\Phi,R,I)$ 
can be factored into a product of one element of $\G(\Delta_1,R,I)$ one element of $Z$ and at most $4(|\Phi^+| - |\Delta_1^+|)-1$ elementary root unipotents $x_\alpha(s)$ of level $I$. \end{cor}
\begin{proof}
The above estimates can be obtained by a careful analysis of the proof of the previous theorem.
Cases $\Phi=\rA_\ell, \rC_\ell$ are immediate.
In the case $\Phi=\rD_\ell$ the proof of Theorem~\ref{thm:BassKolster} implies that
\begin{multline}\nonumber
\G(\Phi,R,I) =  \U(\Sigma_\ell,I) \cdot X_{\alpha_1}(I) \cdot \U(\Sigma_2^-\cap\Delta_1,I) \cdot X_{-\alpha\ssub{\mathrm{max}}}(I) \cdot Z  \cdot \\ \cdot \U(\Sigma_1^-,I) \cdot \U(\Sigma_1,I) \cdot \G(\Delta_1,R,I).
\end{multline}
We can present an element $g$ of $\U(\Sigma_\ell, I)$ as a product of $g_1 \in \U(\Sigma_{\{1,2\}} \cap \Sigma_\ell)$ and $g_2\in \U(\Delta_{\{1,2\}}\cap \Sigma_\ell)$.
An examination of the extended Dynkin diagram of $\rD_\ell$ implies that $g_2$ either centralizes or normalizes all factors of the above decomposition (except the last one) and therefore can be moved to the right until it is consumed by $\G(\Delta_1)$.
On the other hand, $g_1$ is a product of at most $2\ell-3$ elementary unipotents, while the width of $\U(\Sigma_1^\pm, I)$ and $\U(\Sigma_2^-\cap\Delta_1)$ in elementary unipotents does not exceed $2\ell-2$ and $2\ell-4$, respectively.
Summing up these upper bounds we obtain
$$(2\ell-3) + 1 + (2\ell - 4) + 1 + 2\cdot (2\ell - 2) = 8\ell - 9 = 4(|\rD_\ell| - |\rD_{\ell-1}|) - 1.$$

@@

\begin{proof}[Proof of Theorem~\ref{thm:SL2width}]
Consider the case $\Phi=\rA_\ell$, $\sr(I)\leq 2$. First of all, notice that we can improve the estimate of the number of factors involved in Bass---Kolster decomposition.
Indeed, when performing the first step of the proof of Theorem~\ref{thm:BassKolster} it suffices to make only $2$ additions of $v_{\ell+1}$ (e.\,g. to $v_{1}$ and $v_2$) to make the first $\ell$ entries of $v'$ form a unimodular column.
In particular, $\G(\Phi, R, I)$ can be presented as a product of one element of $\G(\Delta_1, R, I)$, one element of $Z$ and $3\ell+1$ root unipotents $x_\alpha(s)$, $s\in I$.
The latter elements are contained in a product of $3\ell + 1$ copies of $\SL(2, R, I)$. Now the statement of the theorem follows by induction on $\ell$.

The proof in the case $\Phi=\rC_\ell$ is similar (notice that we use the exceptional isomorphism $\SL(2, R)\cong \Sp(2, R)$).
\end{proof}

### 2015-08-26 congrsubgr-boundgen.tex
Slight changes in the notation; further reorganization

\begin{lemma}\label{lemma:srRI1}
Let $\Phi$ be a root system, $I$ be an ideal of $R$ such that $\sr(I)=1$. Then the relative elementary Chevalley group admits Gauss decomposition
\[ \E(\Phi,R,I) = \Hh(\Phi,R,I) \U(\Phi,I)\U^-(\Phi,I)\U(\Phi,I). \]
As a corollary, the width of $\E(\Phi,R,I)$ with respect to the generating set $\mathcal{Z}$ does not exceed $3\left|\Phi^+\right|+2\rk\Phi-1$.
\end{lemma}
\begin{proof}
Let $A=\begin{psmallmatrix}a & b \\ c & d\end{psmallmatrix}\in\SL(2,R,I)$ be an element of $\SL(2, R, I)$.
The first column of $A$ is $I$-unimodular, therefore there exists $z\in I$ such that $a+cz\in R^*$.
Multiplying $A$ on the left by $x_{12}(z)$ we get matrix $A'=x_{12}(z)\cdot A=\begin{psmallmatrix}a' & b' \\ c & d\end{psmallmatrix}$ with invertible top-left corner element $a'$.
After multiplying $A'$ on the left by $x_{21}(-c/a')$ and on the right by $x_{12}(-b'/a')$ we get a diagonal matrix. 
Thus we obtain relative Gauss decomposition of $A$
\begin{equation}\nonumber
A=x_{12}(-z)\cdot x_{21}(c/a')\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{12}(b'/a')=x_{12}(-z)\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{21}(y) \cdot x_{12}(b'/a'),
\end{equation}
where $\varepsilon\in 1+I$ and $y\in I$. Theorem~\ref{thm:Gauss} implies that the Gauss decomposition of length $3$ hold for every relative elementary Chevalley group.

To obtain the width estimates fix an element $g\in\E(\Phi,R,I)$ and decompose it into $g=u_1 h v_2 u_3$, where $h\in\Hh(\Phi,R,I)$, $u_1,u_3\in\U(\Phi,I)$, $v_2\in\U^-(\Phi,I)$. Write $h=\prod_{i=1}^\ell h_{\alpha_i}(\varepsilon_i)$, $\varepsilon\in1+I$. Each $h_{\alpha_i}(\varepsilon_i)$ decomposes into $h_{\alpha_i}(\varepsilon_i) = x_{\alpha_i}(*) z_{-\alpha_i}(*,*) x_{-\alpha_i}(*)$ (see Lemma~\ref{lemma:rel-tor-elementary}), and since the torus normalizes each of $\X_\alpha(I)$ (see formula~\ref{rel:R6}), we have a decomposition
\[ g\in\U(\Phi,I)\cdot\prod_{\i=1}^\ell\bigl(x_{\alpha_i}(*)z_{-\alpha_i}(*,*)\bigr)\cdot \U^-(\Phi,I) \U(\Phi,I), \]
and the estimate follows.
%
%
%Direct calculation shows that
%\begin{multline*}
%\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix} =
%\begin{pmatrix} 1 & -1 \\ 0 & 1 \end{pmatrix}
%\begin{pmatrix} 1 & 0 \\ 1-\varepsilon & 1 \end{pmatrix}
%\begin{pmatrix} 1 & 1/\varepsilon \\ 0 & 1 \end{pmatrix}
%\begin{pmatrix} 1 & 0 \\ \varepsilon^2-\varepsilon & 1 \end{pmatrix} =\\=
%\begin{pmatrix} 1 & -1 \\ 0 & 1 \end{pmatrix}
%\begin{pmatrix} 1 & 0 \\ 1-\varepsilon & 1 \end{pmatrix}
%\begin{pmatrix} 1 & 1+z \\ 0 & 1 \end{pmatrix}
%\begin{pmatrix} 1 & 1/\varepsilon-1-z \\ 0 & 1 \end{pmatrix}
%\begin{pmatrix} 1 & 0 \\ \varepsilon^2-\varepsilon & 1 \end{pmatrix},
%\end{multline*}
%therefore
%\[ A=z_{21}(1-\varepsilon,-1-z)\cdot x_{12}(1/\varepsilon-1-z)\cdot x_{21}(\varepsilon^2-\varepsilon+y)\cdot x_{12}(b'/a'), \]
%which simultaneously implies both statements of the lemma.
\end{proof}

@@

%Now it is easy to derive bounded generation property for $\G(\Phi,\mathcal{O}_S,I)$ from the above statements.
\begin{lemma}\ref{lemma:width-dedekind}
Let $\Phi$ be an irreducible classical root system of rank $\ell \geqslant 2$.
If $k$ has a real embedding, then $\G(\Phi,\mathcal{O}_S,I)$ has finite width with respect to the generating set $\mathcal{Z}(\Sigma_\ell)$.
\end{lemma}
\begin{proof}
First of all, notice that $\asr(I) \leq \asr(\mathcal{O}_S) \leq 2$. 
Inductively applying Theorem~\ref{thm:BassKolster} and using exceptional isomorphisms for small-rank groups we can present any element of $G=\G(\Phi, \mathcal{O}_S, I)$ as a product of a finite number of generators from $\mathcal{Z}(\Sigma_\ell)$ and one element of $\G(\{\alpha_\ell, -\alpha_\ell\} \mathcal{O}_S, I)\cong\SL(2,\mathcal{O}_S,I)$.
Consequently, to prove the statement of the lemma it suffices to express every element $g = \begin{psmallmatrix}1+a & b \\ c & 1+d \end{psmallmatrix} \in G_0$ as a product of a finite number of generators contained in some rank $2$ subgroup of $G$ containing $G_0$.

@@

>>>>

\begin{lemma}\label{lemma:srRI1_width} In the assumptions of Lemma~\ref{lemma:srRI1} the width of $\E(\Phi,R,I)$ with respect to $\mathcal{Z}(\Pi)$ does not exceed $3\left|\Phi^+\right|+2\rk\Phi-1$. \end{lemma}
\begin{proof}
Take an element $g\in\E(\Phi,R,I)$ and decompose it into $g=u_1 h v_2 u_3$, where $h\in\Hh(\Phi,R,I)$, $u_1,u_3\in\U(\Phi,I)$, $v_2\in\U(\Phi^-,I)$. 
Write $h=\prod_{i=1}^\ell h_{\alpha_i}(\varepsilon_i)$, $\varepsilon\in1+I$. 
Each $h_{\alpha_i}(\varepsilon_i)$ decomposes into $h_{\alpha_i}(\varepsilon_i) = x_{\alpha_i}(*) z_{-\alpha_i}(*,*) x_{-\alpha_i}(*)$ 
 (see Lemma~\ref{lemma:rel-tor-elementary}), and since the torus normalizes each of $\X_\alpha(I)$ (see formula~\ref{rel:R6}), we have a decomposition
\[ g\in\U(\Phi,I)\cdot\prod_{\i=1}^\ell\bigl(x_{\alpha_i}(*)z_{-\alpha_i}(*,*)\bigr)\cdot \U^-(\Phi,I) \U(\Phi,I), \]
and the estimate follows. \end{proof}

@@

%Now it is easy to derive bounded generation property for $\G(\Phi,\mathcal{O}_S,I)$ from the above statements.
\begin{lemma}\label{lemma:width-dedekind}
Let $\Phi$ be an irreducible classical root system of rank $\ell \geqslant 2$.
If $k$ has a real embedding, then $\G(\Phi,\mathcal{O}_S,I)$ has finite width with respect to the generating set $\mathcal{Z}(\Sigma_\ell)$.
\end{lemma}
\begin{proof}
First of all, notice that $\asr(I) \leq \asr(\mathcal{O}_S) \leq 2$. 
Inductively applying Theorem~\ref{thm:BassKolster} and using exceptional isomorphisms for small-rank groups we can present any element 
of $G=\G(\Phi, \mathcal{O}_S, I)$ as a product of a finite number of generators from $\mathcal{Z}(\Sigma_\ell)$ and one element of 
$\G(\{\alpha_\ell, -\alpha_\ell\}, \mathcal{O}_S, I)\cong\SL(2,\mathcal{O}_S,I)$.
Consequently, to prove the statement of the lemma it suffices to express every element 
$g = \begin{psmallmatrix}1+a & b \\ c & 1+d \end{psmallmatrix} \in G_0$
as a product of a finite number of generators contained in some rank $2$ subgroup of $G$ containing $G_0$.

@@

\begin{proof}[Proof of Theorem~\ref{thm:width}] The first statement of the theorem is contained in Lemma~\ref{lemma:srRI1_width}.
The second and third statement follow from Lemmas~\ref{lemma:Z1p}--\ref{lemma:width-dedekind}. \end{proof}

### 2015-08-26 gauss.tex
Slight changes in the notation; further reorganization

\begin{thm}\label{thm:Gauss}
Let $\Phi$ be a reduced irreducible root system of rank $\ell$ and let $\Delta_1, \Delta_\ell$ be
two reductive subsystems of $\Phi$ corresponding to the endnodes of the Dynkin diagram of $\Phi$.
Suppose that both relative elementary subgroups $\E(\Delta_i, R, I)$, $i=1,\ell$ admit triangular decomposition with $N$ triangular factors:
\[ \E(\Delta_i, R, I) = \Hh(\Delta_i, R, I) \cdot \U(\Delta^+_i, I) \cdot \U(\Delta^-_i, I) \cdot \ldots \cdot \U(\Delta^\pm_i, I),\quad i=1,\ell. \]
Then the group $\E(\Phi, R, I)$ admits the decomposition with the same number of factors:
\[ \E(\Phi, R, I) = \Hh(\Phi,R,I) \cdot \U(\Phi^+,I) \cdot \U(\Phi^-, I) \cdot \ldots \cdot \U(\Phi^\pm,I). \]
\end{thm}
\begin{proof}
Denote by $Y$ the product of subgroups in the right-hand side of the above equality.
To show that $Y=\E(\Phi,R,I)$ it suffices to check that
\begin{enumerate}
\item $Y$ is normalized by $\E(\Phi,R)$, i.\,e. $Y^{\E(\Phi,R)}\subseteq Y$;
\item there exists a set $X$ generating $\E(\Phi,R,I)$ as a \emph{normal} subgroup of $\E(\Phi,R)$ such that $XY\subseteq Y$.
\end{enumerate}
To prove the first assertion it suffices to show that $Y^{x_\alpha(\xi)} \subseteq Y$ for any $\alpha\in \pm \Pi$, $\xi\in R$.
Fix a root $\alpha\in\pm\Pi$. Clearly, $\alpha \in \Delta_i$ for $i=1$ or $i=\ell$ and we can expand $Y$ as follows:
\[ Y=\Hh(\Phi,R,I) \cdot \U(\Delta_i^+,I) \cdot \ldots \cdot \U^\pm(\Delta_i,I) \cdot \U(\Sigma_i^+,I) \cdot \ldots \cdot \U(\Sigma_i^\pm,I). \]
For every  $h \in \Hh(\Phi, R, I)$ one has for some $s\in I$
\[ x_\alpha(\xi)\cdot h = h\cdot x_\alpha((1+s)\xi)=h\cdot x_\alpha(\xi)\,x_\alpha(s\xi). \]

@@

>>>>

The proof of Gauss decomposition presented below is similar to the absolute case (cf.~\cite[Theorem~5.1]{Sm12}).

\begin{prop}\label{thm:Gauss}
Let $\Phi$ be a reduced irreducible root system of rank $\ell$ and let $\Delta_1, \Delta_\ell$ be
two reductive subsystems of $\Phi$ corresponding to the endnodes of the Dynkin diagram of $\Phi$.
Suppose that both relative elementary subgroups $\E(\Delta_i, R, I)$, $i=1,\ell$ admit triangular decomposition with $N$ triangular factors:
\[ \E(\Delta_i, R, I) = \Hh(\Delta_i, R, I) \cdot \U(\Delta^+_i, I) \cdot \U(\Delta^-_i, I) \cdot \ldots \cdot \U(\Delta^\pm_i, I),\quad i=1,\ell. \]
Then the group $\E(\Phi, R, I)$ admits the decomposition with the same number of factors:
\[ \E(\Phi, R, I) = \Hh(\Phi,R,I) \cdot \U(\Phi^+,I) \cdot \U(\Phi^-, I) \cdot \ldots \cdot \U(\Phi^\pm,I). \]
\end{prop}
\begin{proof}
Denote by $Y$ the product of subgroups in the right-hand side of the above equality.
To show that $Y=\E(\Phi,R,I)$ it suffices to check that
\begin{enumerate}
\item $Y$ is normalized by $\E(\Phi,R)$, i.\,e. $Y^{\E(\Phi,R)}\subseteq Y$;
\item there exists a set $X$ generating $\E(\Phi,R,I)$ as a \emph{normal} subgroup of $\E(\Phi,R)$ such that $XY\subseteq Y$.
\end{enumerate}
To prove the first assertion it suffices to show that $Y^{x_\alpha(\xi)} \subseteq Y$ for any $\alpha\in \pm \Pi$, $\xi\in R$.
Fix a root $\alpha\in\pm\Pi$. Clearly, $\alpha \in \Delta_i$ for $i=1$ or $i=\ell$ and we can expand $Y$ as follows:
\[ Y=\Hh(\Phi,R,I) \cdot \U(\Delta_i^+,I) \cdot \ldots \cdot \U^\pm(\Delta_i,I) \cdot \U(\Sigma_i^+,I) \cdot \ldots \cdot \U(\Sigma_i^\pm,I). \]
For every  $h \in \Hh(\Phi, R, I)$ one has for some $s\in I$
\[ x_\alpha(\xi)\cdot h = h\cdot x_\alpha((1+s)\xi)=h\cdot x_\alpha(\xi)\,x_\alpha(s\xi). \]

@@

\begin{lemma}\label{lemma:srRI1}
Let $\Phi$ be a root system, let $I$ be an ideal of arbitrary commutative ring $R$ such that $\sr(I)=1$.
Then the relative elementary Chevalley group admits Gauss decomposition
\[ \E(\Phi,R,I) = \Hh(\Phi,R,I) \cdot \U(\Phi,I) \cdot \U(\Phi^-,I) \cdot \U(\Phi,I). \]
\end{lemma}
\begin{proof}
In view of the above proposition it suffices to prove that Gauss decomposition of length $3$ holds for $\Phi=\rA_1$.

Let $A=\begin{psmallmatrix}a & b \\ c & d\end{psmallmatrix}\in\SL(2,R,I)$ be an element of $\SL(2, R, I)$.
The first column of $A$ is $I$-unimodular, therefore there exists $z\in I$ such that $a+cz\in R^*$.
Multiplying $A$ on the left by $x_{12}(z)$ we get matrix $A'=x_{12}(z)\cdot A=\begin{psmallmatrix}a' & b' \\ c & d\end{psmallmatrix}$ with invertible top-left corner element $a'$.
After multiplying $A'$ on the left by $x_{21}(-c/a')$ and on the right by $x_{12}(-b'/a')$ we get a diagonal matrix. 
Thus we obtain relative Gauss decomposition of $A$
\begin{equation}\nonumber
A=x_{12}(-z)\cdot x_{21}(c/a')\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{12}(b'/a')=x_{12}(-z)\cdot
\begin{pmatrix} \varepsilon & 0 \\ 0 & 1/\varepsilon \end{pmatrix}
\cdot x_{21}(y) \cdot x_{12}(b'/a'),
\end{equation}
where $\varepsilon\in 1+I$ and $y\in I$. \end{proof}

### 2015-08-26 paper.tex
Slight changes in the notation; further reorganization

Let $G$ be an abstract group with a generating set $X \subset G$. 
We denote by $W(G, X)$ the \emph{width of $G$ with respect to $X$}, i.\,e. the smallest natural number $N$ such that every element of $G$ is a product of at most $N$ elements of $X$.
The problem of finding an estimate for the width of $\G(\Phi, R)$ in terms of elementary generators $x_\alpha(\xi)$ has been extensively studied in a number of works (see e.\,g. \cite{CK83, Ka82, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}).
Summarizing results of these works, one can hope that such width is finite only for rather narrow classes of rings of number-theoretic nature, e.\,g. localizations of Dedeking rings of arithmetic type.
For example, W.~van der Kallen has shown that the width of $\SL_3(R)$ is infinite already for a Euclidean ring $R=\mathbb{C}[t]$.

On the other hand, much less is known about the width of relative groups $\G(\Phi, R, I)$. 
Set $z_\alpha(s, \xi) = x_{\alpha}(s)^{x_{-\alpha}(\xi)}$, $s\in I$, $\xi\in R$.
It is a classical result of J.~Tits and L.~Vaserstein (see e.\,g. Theorem~\cite[Theorem~2]{Va86}) that the relative elementary group $\E(\Phi, R, I)$ is generated by
the set of elements $\mathcal{Z} = \{ z_\alpha(s,\xi) \mid \alpha \in \Phi,\ s\in I,\ \xi \in R \}.$
Recently, A.~Stepanov has shown that, in fact, a more economical set of generators can be chosen for $\E(\Phi, R, I)$ (see~\cite[Theorem~3.4]{S}).
\begin{thm}
Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$ and let $S$ be any parabolic subset of roots of $\Phi$ with the special part $\Sigma_S$ (see section~\ref{sec:rootsys}).
Denote by $\mathcal{Z}(\Sigma_S)$ the subset of $\mathcal{Z}$ consisting of the following two families of elements:
\begin{itemize}
\item $x_{\alpha}(s)$, where $s\in I$, $\alpha\in\Phi$;
\item $z_\alpha(s,\xi)$, where $s\in I$, $\xi\in R$, $\alpha\in\Sigma_S$.
\end{itemize}
Then $\E(\Phi, R, I)$ is generated (as an abstract group) by $\mathcal{Z}(\Sigma_S)$.
\end{thm}

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of $\G(\Phi, R, I)$ in terms of $\mathcal{Z}$.
\begin{thm} Let $\Phi$ be an irreducible root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $W(\E(\Phi,R,I),\mathcal{Z})\leqslant 3|\Phi^+|+2\rk(\Phi)-1$;
\item Let $p$ be a prime integer, set $R = \mathbb{Z}[\sfrac{1}{p}]$ and assume $\Phi$ is classical. Then under the assumption of the Generalized Riemann Hypothesis there is an estimaete  $W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 1$;
\item Let $\mathcal{O}_S$ be a Dedekind ring of arithmetic type embedded into some global field $k$ having a real embedding. Assume that $\Phi$ is classical of rank $\geqslant2$, then $W(\G(\Phi,\mathcal{O}_S,I), \mathcal{Z}(\Sigma_\ell))$ is finite.
\end{enumerate}
\end{thm}
\begin{proof}

%In the proof of the above theorem we use Bass---Kolster decomposition to reduce to the case $\Phi=\rA_1$ which then can be settled directly.
%Notice that the particular case $I=R$ of the above theorem was demonstrated in~\cite{VavSmSuUnitrEng} (see also \cite{VseUnitrZ1p}).
...
\end{proof}
The particular case $\Phi = \rA_\ell$, $R = I = \mathbb{Z}$ of the above theorem is similar in spirit to a recent theorem of U.~Hadad (see ~\cite[Theorem~1.6]{Ha12}).

@@

In the linear and symplectic case the above results can be extended to even larger class of rings (which e.\,g. includes all Dedekind rings).
\begin{thm}\label{thm:SL2width}
Assume that one of the following is true:
\begin{itemize} 
\item $I$ is an ideal of a (not neccessarily commutative) ring $R$ such that $\sr(I)\leq 2$, $\Phi=\rA_\ell$, $l\geq 2$, $N=3|\Phi^+|+\rk(\Phi) - 3$.
\item $I$ is an ideal of a commutative ring $R$ such that $\sr(I)\leq 3$, $\Phi=\rC_\ell$, $l\geq 2$, $N=3|\Phi^+|+2\rk(\Phi) - 4$
\end{itemize}
Then $\G(\Phi, R, I)$ can be presented as a product of at most $N$ copies of its subgroups isomorphic to $\G(\rA_1, R, I)$. \end{thm}

@@

We also define $h_\alpha(\varepsilon)=w_\alpha(\varepsilon)w(-1)$. Then
\begin{equation}\label{rel:R6}
h_\alpha(\varepsilon)x_\beta(\xi)h_\alpha(\varepsilon)^{-1} = x_\beta\left(\varepsilon^{\langle\beta,\alpha\rangle}\xi\right),\quad \varepsilon\in R^*,\ \xi\in R.
\end{equation}

>>>>

Let $G$ be an abstract group with a generating set $X \subset G$. 
We denote by $W(G, X)$ the \emph{width of $G$ with respect to $X$}, i.\,e. the smallest natural number $N$ such that every element of $G$ is a product of at most $N$ elements of $X$.
The problem of finding an estimate for the width of $\G(\Phi, R)$ in terms of elementary generators $x_\alpha(\xi)$ has been extensively studied in a number of works (see e.\,g. \cite{CK83, Ka82, Tavgen91, Mo07, VseUnitrZ1p, VavSmSuUnitrEng}).
Summarizing results of these works, one can hope that such width is finite only for rather narrow classes of rings of number-theoretic nature, e.\,g. localizations of Dedeking rings of arithmetic type.
For example, W.~van der Kallen has shown that the width of $\SL(3,R)$ is infinite already for a Euclidean ring $R=\mathbb{C}[t]$.

On the other hand, much less is known about the width of the relative group $\G(\Phi, R, I)$. 
Set $z_\alpha(s, \xi) = x_{\alpha}(s)^{x_{-\alpha}(\xi)}$, $s\in I$, $\xi\in R$.
It is a classical result of J.~Tits and L.~Vaserstein (see e.\,g.~\cite[Theorem~2]{Va86}) that the relative elementary group $\E(\Phi, R, I)$ is generated by
the set of elements $\mathcal{Z} = \{ z_\alpha(s,\xi) \mid \alpha \in \Phi,\ s\in I,\ \xi \in R \}.$

In fact, one can choose a smaller set of generators for $\E(\Phi, R, I)$. 
Let $\Sigma$ be any special subset of $\Phi$, (e.\,g. the set of simple roots $\Pi$ or the special part of $\Sigma_S$ of some parabolic subset of roots $S$, see section~\ref{sec:rootsys}).
Set $$\mathcal{X} = \{ x_\alpha(s) \mid \alpha\in \Phi,\ s\in I\};\ \ \mathcal{Z}(\Sigma) = \{ z_\alpha(s, \xi) \mid s \in I,\ \xi \in R, \alpha\in \Sigma \} \cup \mathcal{X}.$$

Now we are ready to state a recent theorem of A.~Stepanov (see~\cite[Theorem~3.4]{S}).  
\begin{thm}\label{thm:Stepanov} Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$ and let $S \subset \Phi$ be any parabolic subset of roots.
Then $\E(\Phi, R, I)$ is generated as an abstract group by $\mathcal{Z}(\Sigma_S)$. \end{thm}

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of $\G(\Phi, R, I)$ in terms of certain subsets of $\mathcal{Z}$.
\begin{thm}\label{thm:width} Let $\Phi$ be an irreducible root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $W(\E(\Phi,R,I),\mathcal{Z}(\Pi))\leqslant 3|\Phi^+|+2\rk(\Phi)-1$;
\item Let $p$ be a prime integer, set $R = \mathbb{Z}[\sfrac{1}{p}]$ and assume $\Phi$ is classical. Then under the assumption of the Generalized Riemann Hypothesis one has
 $$W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 1;$$
\item Let $\mathcal{O}_S$ be a Dedekind ring of arithmetic type embedded into some global field $k$ having a real embedding. 
Assume $\Phi$ is classical of rank $\geqslant2$, then $W(\G(\Phi,\mathcal{O}_S,I), \mathcal{Z}(\Sigma_\ell))$ is finite.
\end{enumerate}
\end{thm}
The particular case $\Phi = \rA_\ell$, $R = I = \mathbb{Z}$ of the above theorem is similar in spirit to a recent theorem of U.~Hadad (see ~\cite[Theorem~1.6]{Ha12}).

@@

In the linear and symplectic case the above results can be extended to even larger class of rings (which e.\,g. includes all Dedekind rings).
\begin{thm}\label{thm:SL2width}
Assume that one of the following is true:
\begin{itemize} 
\item $I$ is an ideal such that $\sr(I)\leq 2$, $\Phi=\rA_\ell$, $l\geq 2$, $N=3|\Phi^+|+\rk(\Phi) - 3$.
\item $I$ is an ideal such that $\sr(I)\leq 3$, $\Phi=\rC_\ell$, $l\geq 2$, $N=3|\Phi^+|+2\rk(\Phi) - 4$
\end{itemize}
Then $\G(\Phi, R, I)$ can be presented as a product of at most $N$ copies of its subgroups isomorphic to $\G(\rA_1, R, I)$. \end{thm}

@@

Define semisimple root elements $h_\alpha(\varepsilon)$, $\alpha\in\Phi$, $\varepsilon\in R^*$ as $h_\alpha(\varepsilon)=w_\alpha(\varepsilon)w_\alpha(-1)$. Then
\begin{equation}\label{rel:R6}
h_\alpha(\varepsilon)x_\beta(\xi)h_\alpha(\varepsilon)^{-1} = x_\beta\left(\varepsilon^{\langle\beta,\alpha\rangle}\xi\right),\quad \varepsilon\in R^*,\ \xi\in R.
\end{equation}
